---
title: "Code for Metabolites Dataset"
author: "Jingyu Liang"
date: "2025-01-23"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(ggplot2)
library(magrittr)
library(writexl)
library(reshape2)
```

# data
## metabolites data
**metabolites_clear** is the samples with both post and pre. We deleted the single sample.
**metabolites_diff** is the difference between post and pre.
```{r}
# read the data
metabolites <- read_excel("for_code_UNTX-07-23PHML+ DATA TABLES.xlsx", sheet = "Log Transformed Data")
prepost <- read_excel("for_code_UNTX-07-23PHML+ DATA TABLES.xlsx", sheet = "prepost")
request <- read_excel("for_code_Request analysis_FFHwang 0520.xlsx")
metabolites <- as.data.frame(metabolites)
prepost <- as.data.frame(prepost)
request <- as.data.frame(request)

# request change col name
request %<>% rename(Time = `Time (1=pre; 2=post)`)

# request change the format of dates
request$`Sampled date` <- format(as.Date(request$`Sampled date`), "%m%d%y")
```

get the **metabolites_clear** and **metabolites_diff**
```{r}
# Identify Sample IDs that have both 1 and 2 in request
valid_ids <- request %>%
  group_by(`Sample ID`) %>%
  summarise(unique_times = n_distinct(Time))%>%
  filter(unique_times > 1) %>% # Keeps only rows where unique_times > 1
  pull(`Sample ID`) # Extracts only the Sample ID column as a vector

# drop Sample ID only has 1 or 2 in request_clear
request_clear <- request %>%
  filter(`Sample ID` %in% valid_ids) %>%
  arrange(`Sample ID`,Time)

# merge Sample ID and Sampled date for left_join in request_clear
request_clear <- request_clear %>% 
  mutate(CLIENT_SAMPLE_ID2 = paste(`Sample ID`, `Sampled date`, sep = " "))

# separate the ID column into 2
# prepost <- prepost %>%
#   separate(col=`CLIENT_SAMPLE_ID2`,
#            into = c("Sample ID", "Sampled date"),
#            sep = " ",
#            fill = "right")
# prepost %<>% arrange(`Sample ID`)


# left join the pre and post from request analysis
prepost_clear <- prepost %>% left_join(request_clear[,-c(5:6)], by = "CLIENT_SAMPLE_ID2") %>%
  arrange(`Sample ID`,Time)

# prepost_clear: drop those single Sample ID
prepost_clear <- na.omit(prepost_clear)

# metabolites: change to numeric
metabolites[,-1] <- apply(metabolites[,-1], 2, as.numeric) 

# left_join metabolites
metabolites_clear <- prepost_clear %>% 
  left_join(metabolites, by = "PARENT_SAMPLE_NAME") %>%
  select(-CLIENT_SAMPLE_ID2,-`Sampled date`) %>%
  arrange(`Sample ID`,Time)

# divide metabolites_clear into pre and post
pre <- metabolites_clear %>% filter(Time==1) %>% arrange(`Sample ID`)
post <- metabolites_clear %>% filter(Time==2) %>% arrange(`Sample ID`)

# calculate the difference
metabolites_diff <- metabolites_clear %>%
  group_by(`Sample ID`) %>%
  summarise(across(where(is.numeric), ~ .[Time==2]-.[Time==1]))%>%
  select(-Time)

metabolites_diff<- as.data.frame(metabolites_diff)
```

**metabolites_new** has both pairs and single sample
```{r}
# merge Sample ID and Sampled date for left_join in request
request_new <- request %>% 
  mutate(CLIENT_SAMPLE_ID2 = paste(`Sample ID`, `Sampled date`, sep = " "))

# separate the ID column into 2
# prepost <- prepost %>%
#   separate(col=`CLIENT_SAMPLE_ID2`,
#            into = c("Sample ID", "Sampled date"),
#            sep = " ",
#            fill = "right")
# prepost %<>% arrange(`Sample ID`)


# left join the pre and post from request analysis
prepost_new <- prepost %>% left_join(request_new[,-c(5:6)], by = "CLIENT_SAMPLE_ID2") %>%
  arrange(`Sample ID`,Time)

# prepost_new: drop na
prepost_new <- na.omit(prepost_new)

# left_join metabolites
metabolites_new <- prepost_new %>% 
  left_join(metabolites, by = "PARENT_SAMPLE_NAME") %>%
  select(-CLIENT_SAMPLE_ID2,-`Sampled date`) %>%
  arrange(`Sample ID`,Time)
```

## demographic information dataset
```{r}
demo <- read_excel("Demographic FF 0228.xlsx")
demo$ID <- as.numeric(demo$ID)
demo <- demo %>% 
  mutate(Gender = ifelse(demo$Gender=="F", 1, 0)) %>%
  rename(Female = Gender) %>%
  mutate(`Current smoker` = ifelse(demo$`Current smoker`=="Y", 1, 0)) %>%
  mutate(`Former smoker` = ifelse(demo$`Former smoker`=="Y", 1, 0)) %>%
  rename(`Sample ID` = ID)

```


# Step 1: Paired t-test and Bonferroni correction to clean the data 
(to find metabolites that have significant difference before and after fire)

**metabolites_sig** is gained from metabolites_diff, representing the difference values of metabolites that are significant in paired t test.
**metabolites_sig_prepost** is gained from metabolites_clear, representing the pairs (post and pre values) of metabolites that are significant in paired t-test.
**metabolites_new_sig** is gained from metabolites_new that has both paired and single sample. 
```{r}
# Select only numeric columns for analysis
numeric_cols <- names(metabolites_clear)[sapply(metabolites_clear, is.numeric)]
numeric_cols <- setdiff(numeric_cols, "Time")  # Remove "Time" column: find all rows in numeric_cols that aren't in "Time"

# Perform paired t-test for each variable
t_test_results <- lapply(numeric_cols, function(var) {
  t_test <- t.test(pre[[var]], post[[var]], paired = TRUE)
  return(c(var, t_test$statistic, t_test$p.value))
})

# Convert list to dataframe
t_test_df <- as.data.frame(do.call(rbind, t_test_results))
colnames(t_test_df) <- c("CHEM_ID", "t_statistic", "p_value")

# Convert numeric columns to proper format
t_test_df$t_statistic <- as.numeric(t_test_df$t_statistic)
t_test_df$p_value <- as.numeric(t_test_df$p_value)

# Number of tests
num_tests <- nrow(t_test_df)

# Bonferroni-corrected p-value threshold (adjusted significance level)
alpha <- 0.05
bonferroni_threshold <- alpha / num_tests

# Apply Bonferroni correction
t_test_df$adjusted_p_value <- p.adjust(t_test_df$p_value, method = "bonferroni")
t_test_df$significant <- t_test_df$adjusted_p_value < alpha

# View final results with correction
print(t_test_df)

# filter the significant chemicals
t_test_clear <- t_test_df %>%
  filter(significant==TRUE)

# get the CHEM_ID of significant Chemicals
significant_CHEM <- t_test_clear$CHEM_ID
```

```{r}
# clear the metabolites
metabolites_sig <- metabolites_diff %>%
  select(`Sample ID`, all_of(significant_CHEM))

# metabolites with pre and post
metabolites_sig_prepost <- metabolites_clear %>%
  select(`Sample ID`, Time, Group, all_of(significant_CHEM))

# all paired and single metabolites
metabolites_new_sig <- metabolites_new %>%
  select(PARENT_SAMPLE_NAME, `Sample ID`, Time, Group, all_of(significant_CHEM))
# write_xlsx(metabolites_sig, "metabolites_sig.xlsx")
```

# summary statistics
```{r}
summary_table <- apply(metabolites[-1], 2, function(x) {
  c(
    Mean = mean(x, na.rm = TRUE),
    Median = median(x, na.rm = TRUE),
    Q1 = quantile(x, 0.25, na.rm = TRUE),
    Q3 = quantile(x, 0.75, na.rm = TRUE),
    Min = min(x, na.rm = TRUE),
    Max = max(x, na.rm = TRUE)
  )
})

# Convert to a data frame for better readability
summary_table <- as.data.frame(t(summary_table))
summary_table <- round(summary_table, 2)
print(summary_table)
```
# Normality Check
## Histograms
```{r}

# draw the histograms
par(mfrow=c(6,5))
for (col in names(sample)){
  sample_data <- sample[[col]]
  hist(sample_data, main=paste(col), xlab = "values", col="lightblue", border="black")
  # result <- shapiro.test(sample_data)
}

```


## Shapiro-Wilk Test
```{r}
# Identify columns with variation
# valid_columns <- apply(metabolites_diff[, -1], 2, function(x) length(unique(x)) > 1)

# Run Shapiro-Wilk test only on columns with variation
shapiro_results <- apply(metabolites_diff[, -1], 2, function(x) shapiro.test(x)$p.value)

# Display normal columns (p > 0.05)
normal_columns <- names(shapiro_results[shapiro_results > 0.05])
print(normal_columns)

# Display not normal columns (p < 0.05)
not_normal_columns <- names(shapiro_results[shapiro_results < 0.05])
print(not_normal_columns)
```

# Box Plot
```{r}
# organize the data to make box plot
reshaped_cols <- colnames(metabolites_sig_prepost)[4:34]
prepost_long <- melt(metabolites_sig_prepost, id.vars = c("Sample ID","Time","Group"), measure.vars = reshaped_cols, variable.name = "CHEM_ID", value.name="Values")

prepost_long <- prepost_long %>%
  group_by(CHEM_ID) %>%
  mutate(max = max(Values)) %>%
  ungroup() %>%
  arrange(desc(max)) %>%
  mutate(CHEM_ID = factor(CHEM_ID, levels = unique(CHEM_ID)))

# Convert Time to a factor and explicitly set the order
prepost_long$Time <- factor(prepost_long$Time, levels = c("1", "2"), labels = c("pre-fire", "post-fire"))

```

box plot for pre and post fire
```{r}
ggplot(prepost_long, aes(x = CHEM_ID, y = Values, fill = Time)) +
  geom_boxplot(na.rm = TRUE) +
  scale_fill_manual(
    values = c("pre-fire" = "yellow", "post-fire" = "orange"),
    name = "Time"
  )+
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    #panel.background = element_rect(fill = "white", color = "black"),  # White panel with black border
    #plot.background = element_rect(fill = "white", color = NA)  # White outer background
    panel.grid.major = element_blank(), 
	  panel.grid.minor = element_blank(),
	  panel.background = element_blank(), 
    axis.line = element_line(colour = "black"),
    panel.spacing.x = unit(8, "lines")
    )+
  labs(title = "Box Plot of Metabolites with Significant Differences in Paired t-Test ")

```

box plot for CF and VF
```{r}
ggplot(prepost_long, aes(x = CHEM_ID, y = Values, fill = Group)) +
  geom_boxplot(na.rm = TRUE) +
  scale_fill_manual(
    values = c("CF" = "lightblue", "VF" = "lightpink"),
    name = "CF/VF"
  )+
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.major = element_blank(), 
	  panel.grid.minor = element_blank(),
	  panel.background = element_blank(), 
    axis.line = element_line(colour = "black"),
    panel.spacing.x = unit(8, "lines")
    )+
  labs(title = "Box Plot of Metabolites with Significant Differences in Paired t-Test ")

```


# Step 2: Linear Mixed Model (find sig, same as paired t-test)
## build linear mixed models
to find if there is any difference varying by time (pre and post)
```{r}
# library(lme4)
# 
# metabolites_new$Time <- as.factor(metabolites_new$Time)
# metabolites_new$`Sample ID` <- as.factor(metabolites_new$`Sample ID`)

# Example for one metabolite (replace 'Metabolite_X' with actual column name)
# model <- lmer(`209` ~ Time + (1 | `Sample ID`), data = metabolites_new)
# summary(model)

# results <- lapply(names(metabolites_new)[5:ncol(metabolites_new)], 
#                   function(metabolite) {
#                      formula_str <- paste0("`", metabolite, "` ~ Time + (1 | `Sample ID`)")
#                       model <- lmer(as.formula(formula_str), data = metabolites_new)
#                       summary(model)
#                       })
# # example
# formula_str <- paste0("`", "212", "` ~ Time + (1 | `Sample ID`)")

```


## get the p-values etc. from the models
```{r}
# Load necessary libraries
library(lme4)
library(lmerTest)  # To obtain p-values and degrees of freedom
library(broom.mixed)  # For confidence intervals
library(openxlsx)  # To save results
library(dplyr)

metabolites_new$Time <- as.factor(metabolites_new$Time)
metabolites_new$`Sample ID` <- as.factor(metabolites_new$`Sample ID`)

# Function to fit LMM and extract statistics
results_df <- lapply(names(metabolites_new)[5:ncol(metabolites_new)], function(metabolite) {
  tryCatch({
    # Construct formula
    formula_str <- paste0("`", metabolite, "` ~ Time + (1 | `Sample ID`)")
    
    # Fit model using lmerTest::lmer() to get df
    model <- lmerTest::lmer(as.formula(formula_str), data = metabolites_new)
    
    # Print if the model is singular
    # print(isSingular(model))
    
    # Extract summary of fixed effects
    summary_model <- summary(model)
    
    # Extract coefficient matrix and convert to data frame
    fixed_effects <- as.data.frame(summary_model$coefficients)
    
    # Extract correct degrees of freedom (df) for Time effect
    df_time <- fixed_effects["Time2", "df"]  # Correct indexing
    
    # Compute confidence intervals
    ci <- confint(model, method = "Wald")
    
    # Ensure "Time2" effect exists
    time_effect_row <- which(rownames(fixed_effects) == "Time2")
    
    if (length(time_effect_row) > 0) {
      return(data.frame(
        Metabolite = metabolite,
        Estimate = fixed_effects["Time2", "Estimate"],
        Std_Error = fixed_effects["Time2", "Std. Error"],
        t_value = fixed_effects["Time2", "t value"],
        df = df_time,  # Correctly extracted degrees of freedom
        p_value = fixed_effects["Time2", "Pr(>|t|)"],  # Directly extract p-value
        CI_Lower = ci[time_effect_row + 1, 1],  # Lower 95% CI
        CI_Upper = ci[time_effect_row + 1, 2]   # Upper 95% CI
      ))
    } else {
      return(NULL)  # Return NULL if "Time2" effect is missing
    }
  }, error = function(e) {
    return(NULL)  # Ignore errors and continue
  })
})

# Combine results into a single data frame
results_df <- do.call(rbind, results_df)

# Mark the significance level
results_df <- results_df %>%
  mutate(
    significance_original_p = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ NA
    )
  )

# write_xlsx(results_df, "LMM_results.xlsx")
```


## do multiple comparison correction for LMM
alpha = 0.05 means that you expect 5 false positives just by chance.
Multiple testing correction controls for this inflation to ensure that your significant results are truly meaningful.

```{r}
# Apply Bonferroni Correction
results_df$bonferroni_p <- p.adjust(results_df$p_value, method = "bonferroni")

# Apply Benjamini-Hochberg (BH) Correction (FDR)
results_df$bh_p <- p.adjust(results_df$p_value, method = "BH")

# get the significance
results_df <- results_df %>%
  mutate(
    significance_bon_p = case_when(
      bonferroni_p < 0.05 ~ "*",
      TRUE ~ NA
    )
  ) %>%
  mutate(
    significance_bh_p = case_when(
      bh_p < 0.05 ~ "*",
      TRUE ~ NA
    )
  )

# write.xlsx(results_df, file = "LMM_results.xlsx", rowNames = FALSE)

# filter out non-significant rows
results_sig <- na.omit(results_df)

```

## **common_CHEM_ID** the overlapped CHEM_ID with results_sig (LMM) and metabolites_sig (t-test)
**metabo_diff_common** get the overlapped metabolites from metabolites_diff (only difference of pairs)
**metabodiff_demo** metabo_diff_common + demo: overlapped metabolites with demographic information
```{r}
CHEM_ID_t_test <- colnames(metabolites_sig)[-1]
CHEM_ID_LMM <- results_sig$Metabolite

# Find common elements
common_CHEM_ID <- intersect(CHEM_ID_t_test, CHEM_ID_LMM)

common_CHEM_ID

metabo_diff_common <- metabolites_diff %>%
  select(1, all_of(common_CHEM_ID)) %>%
  mutate(`Sample ID` = gsub("FFP01", "", `Sample ID`)) %>%
  mutate(`Sample ID`= as.numeric(`Sample ID`))

metabodiff_demo <- metabo_diff_common %>%
  left_join(demo, metabo_diff_common, by = "Sample ID")


```

# correlation coefficient table
in metabodiff_demo, all firefighters are white, only 2 of them are volunteers, none of them smoking, only one female
**metaodiff_demo_clear** is cleared metabodiff_demo w/o race, CF/VF, smoke, female; and have one data imputed in "Years served as a firefighter"
only "Fire-related activities (times/month)" has two non-significant p-values
```{r}
demo_columns <- c("Age", "BMI", "Years served as a firefighter", 
                  "Non-fire-related activities (times/week)", 
                  "Fire-related activities (times/month)")

# Transform Height and Weight to BMI
metabodiff_demo <- metabodiff_demo %>%
  mutate(BMI = `Weight (kg)`/(`Height (m)`)^2)

metabodiff_demo_clear <- metabodiff_demo %>% 
  select(all_of(c("Sample ID",common_CHEM_ID, demo_columns)))

# input the missing data
library(mice)
# m=1 specifies the number of multiple imputations (since you have only one missing value, one imputation is sufficient).
# method='pmm' specifies the predictive mean matching method.
# seed=500 sets a seed for reproducibility.
imputed_data <- mice(metabodiff_demo_clear, m=1, method='pmm', seed=500)
metabodiff_demo_clear <- complete(imputed_data, 1)

# Generate the correlation coefficient table
correlation_table <- metabodiff_demo_clear %>%
  cor() %>%
  .[common_CHEM_ID, demo_columns]

# t-test and p-values
get_p_values <- function(df, common_CHEM_ID, demo_columns) {
  p_values <- matrix(NA, nrow = length(common_CHEM_ID), ncol = length(demo_columns))
  rownames(p_values) <- common_CHEM_ID
  colnames(p_values) <- demo_columns
  
  for (chem_id in common_CHEM_ID) {
    for (demo_col in demo_columns) {
      group1 <- metabodiff_demo[[chem_id]]
      group2 <- metabodiff_demo[[demo_col]]
      t_test_result <- t.test(group1, group2)
      p_values[chem_id, demo_col] <- t_test_result$p.value
    }
  }
  return(as.data.frame(p_values))
}

# Get p-values for the correlation table
p_values_table <- get_p_values(metabodiff_demo_clear, common_CHEM_ID, demo_columns)

# write.xlsx(p_values_table, file = "corr_p_values_table.xlsx", rowNames = FALSE)
```


# Step 3: Build 2nd model: Linear Mixed Model with all covariates: 
"Age", "BMI", "Years served as a firefighter",
"Non-fire-related activities (times/week)",
"Fire-related activities (times/month)"
Based on the result we got from the above, we have metabolites that have significant difference, common_CHEM_ID, then extract the common_CHEM_ID from the metabolites_new.
```{r}

# combine the demo to metabolites_new (pairs and single)
metabolites_new_common <- metabolites_new %>%
  select(1:4, all_of(common_CHEM_ID)) %>%
  mutate(`Sample ID` = gsub("^([0-9]{1,2}).*", "\\1", `Sample ID`)) %>%
  mutate(`Sample ID` = as.numeric(`Sample ID`))

metabonew_demo <- metabolites_new_common %>%
  left_join(demo, metabolites_new_common, by = "Sample ID")

# Transform Height and Weight to BMI
metabonew_demo <- metabonew_demo %>%
  mutate(BMI = `Weight (kg)`/(`Height (m)`)^2)

metabonew_demo_clear <- metabonew_demo %>% 
  select(all_of(c("Sample ID","Time", "Group", "Current smoker", "Former smoker", common_CHEM_ID, demo_columns)))

# input the missing data
imputed_data <- mice(metabonew_demo_clear, m=1, method='pmm', predictorMatrix = make.predictorMatrix(metabonew_demo_clear))

pred <- make.predictorMatrix(metabonew_demo_clear)
pred[, "Non-fire-related activities (times/week)"] <- 0
pred[, "Fire-related activities (times/month)"] <- 0
pred["Non-fire-related activities (times/week)", ] <- 1
pred["Fire-related activities (times/month)", ] <- 1

imputed_data <- mice(metabonew_demo_clear, m=1, method='pmm', predictorMatrix = pred, seed=500)

metabonew_demo_clear <- complete(imputed_data, 1)

# Function to fit LMM and extract statistics
results_df_2 <- lapply(common_CHEM_ID, function(metabolite) {
  tryCatch({
    # Construct formula with additional covariates
    formula_str <- paste0("`", metabolite, "` ~ Time + Age + BMI + `Years served as a firefighter` + `Non-fire-related activities (times/week)` + `Fire-related activities (times/month)` + Group + `Current smoker` + `Former smoker` + (1 | `Sample ID`)")
    
    # Fit model using lmerTest::lmer() to get df
    model <- lmerTest::lmer(as.formula(formula_str), data = metabonew_demo_clear)
    
    # Extract summary of fixed effects
    summary_model <- summary(model)
    
    # Extract coefficient matrix and convert to data frame
    fixed_effects <- as.data.frame(summary_model$coefficients)
    
    # Compute confidence intervals
    ci <- confint(model, method = "Wald")
    
    # Extract statistics for each covariate
    covariates <- c("Time2", "Age", "BMI", "`Years served as a firefighter`", "`Non-fire-related activities (times/week)`", "`Fire-related activities (times/month)`", "GroupVF", "`Current smoker`", "`Former smoker`")
    results_list <- lapply(covariates, function(covariate) {
      if (covariate %in% rownames(fixed_effects)) {
        return(data.frame(
          Metabolite = metabolite,
          Covariate = covariate,
          Estimate = fixed_effects[covariate, "Estimate"],
          Std_Error = fixed_effects[covariate, "Std. Error"],
          t_value = fixed_effects[covariate, "t value"],
          df = fixed_effects[covariate, "df"],
          p_value = fixed_effects[covariate, "Pr(>|t|)"],
          CI_Lower = ci[covariate, 1],
          CI_Upper = ci[covariate, 2]
        ))
      } else {
        return(NULL)
      }
    })
    
    # Combine results for all covariates
    do.call(rbind, results_list)
  }, error = function(e) {
    return(NULL)  # Ignore errors and continue
  })
})

# Combine results into a single data frame
results_df_2 <- do.call(rbind, results_df_2)

# Mark the significance level
results_df_2 <- results_df_2 %>%
  mutate(
    significance_original_p = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ NA
    )
  )
```


```{r}
# Apply Bonferroni Correction
results_df_2$bonferroni_p <- p.adjust(results_df_2$p_value, method = "bonferroni")

# Apply Benjamini-Hochberg (BH) Correction (FDR)
results_df_2$bh_p <- p.adjust(results_df_2$p_value, method = "BH")

# get the significance
results_df_2 <- results_df_2 %>%
  mutate(
    significance_bon_p = case_when(
      bonferroni_p < 0.05 ~ "*",
      TRUE ~ NA
    )
  ) %>%
  mutate(
    significance_bh_p = case_when(
      bh_p < 0.05 ~ "*",
      TRUE ~ NA
    )
  )

# write.xlsx(results_df_2, file = "LMM_results_2.xlsx", rowNames = FALSE)

# filter out non-significant rows
results_sig_2 <- na.omit(results_df_2)

```
