---
title: "Code for Metabolites Dataset"
author: "Jingyu Liang"
date: "2025-01-23"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(magrittr)
library(writexl)
library(reshape2)
library(mice) # input the missing data
library(ggrepel)
```

# data
## metabolites data
**metabolites_clear** is the samples with both post and pre. We deleted the single sample.
**metabolites_diff** is the difference between post and pre.
```{r}
# read the data
metabolites <- read_excel("for_code_UNTX-07-23PHML+ DATA TABLES.xlsx", sheet = "Log Transformed Data")
prepost <- read_excel("for_code_UNTX-07-23PHML+ DATA TABLES.xlsx", sheet = "prepost")
request <- read_excel("for_code_Request analysis_FFHwang 0520.xlsx")
metabolites <- as.data.frame(metabolites)
prepost <- as.data.frame(prepost)
request <- as.data.frame(request)

# request dataset change col name
request %<>% rename(Time = `Time (1=pre; 2=post)`)

# request dataset change the format of dates
request$`Sampled date` <- format(as.Date(request$`Sampled date`), "%m%d%y")
```

get the **metabolites_clear** and **metabolites_diff**
```{r}
# Identify Sample IDs that have both 1 and 2 in request
valid_ids <- request %>%
  group_by(`Sample ID`) %>%
  summarise(unique_times = n_distinct(Time))%>%
  filter(unique_times > 1) %>% # Keeps only rows where unique_times > 1
  pull(`Sample ID`) # Extracts only the Sample ID column as a vector

# drop Sample ID only has 1 or 2 in request_clear
request_clear <- request %>%
  filter(`Sample ID` %in% valid_ids) %>%
  arrange(`Sample ID`,Time)

# merge Sample ID and Sampled date for left_join in request_clear
request_clear <- request_clear %>% 
  mutate(CLIENT_SAMPLE_ID2 = paste(`Sample ID`, `Sampled date`, sep = " "))

# separate the ID column into 2
# prepost <- prepost %>%
#   separate(col=`CLIENT_SAMPLE_ID2`,
#            into = c("Sample ID", "Sampled date"),
#            sep = " ",
#            fill = "right")
# prepost %<>% arrange(`Sample ID`)


# left join the pre and post from request analysis
prepost_clear <- prepost %>% left_join(request_clear[,-c(5:6)], by = "CLIENT_SAMPLE_ID2") %>%
  arrange(`Sample ID`,Time)

# prepost_clear: drop those single Sample ID
prepost_clear <- na.omit(prepost_clear)

# metabolites: change to numeric
metabolites[,-1] <- apply(metabolites[,-1], 2, as.numeric) 

# left_join metabolites
metabolites_clear <- prepost_clear %>% 
  left_join(metabolites, by = "PARENT_SAMPLE_NAME") %>%
  dplyr::select(-`CLIENT_SAMPLE_ID2`,-`Sampled date`) %>%
  arrange(`Sample ID`,Time)

# divide metabolites_clear into pre and post
pre <- metabolites_clear %>% filter(Time==1) %>% arrange(`Sample ID`)
post <- metabolites_clear %>% filter(Time==2) %>% arrange(`Sample ID`)

# calculate the difference
metabolites_diff <- metabolites_clear %>%
  group_by(`Sample ID`) %>%
  summarise(across(where(is.numeric), ~ .[Time==2]-.[Time==1]))%>%
  dplyr::select(-Time)

metabolites_diff<- as.data.frame(metabolites_diff)
```

**metabolites_new** has *both* pairs and single sample
```{r}
# merge Sample ID and Sampled date for left_join in request
request_new <- request %>% 
  mutate(CLIENT_SAMPLE_ID2 = paste(`Sample ID`, `Sampled date`, sep = " "))

# separate the ID column into 2
# prepost <- prepost %>%
#   separate(col=`CLIENT_SAMPLE_ID2`,
#            into = c("Sample ID", "Sampled date"),
#            sep = " ",
#            fill = "right")
# prepost %<>% arrange(`Sample ID`)


# left join the pre and post from request analysis
prepost_new <- prepost %>% left_join(request_new[,-c(5:6)], by = "CLIENT_SAMPLE_ID2") %>%
  arrange(`Sample ID`,Time)

# prepost_new: drop na
prepost_new <- na.omit(prepost_new)

# left_join metabolites
metabolites_new <- prepost_new %>% 
  left_join(metabolites, by = "PARENT_SAMPLE_NAME") %>%
  dplyr::select(-CLIENT_SAMPLE_ID2,-`Sampled date`) %>%
  arrange(`Sample ID`,Time)
```

## demographic information dataset & impute missing value
demo_all: contains all covariates
demo_select: remove Female, Race,firefighter type, smoke
```{r}
demo <- read_excel("Demographic FF 0228.xlsx")
demo$ID <- as.numeric(demo$ID)
demo <- demo %>% 
  mutate(Gender = ifelse(demo$Gender=="F", 1, 0)) %>%
  rename(Female = Gender) %>%
  mutate(`Current smoker` = ifelse(demo$`Current smoker`=="Y", 1, 0)) %>%
  mutate(`Former smoker` = ifelse(demo$`Former smoker`=="Y", 1, 0)) %>%
  rename(`Sample ID` = ID) %>%
  mutate(BMI = `Weight (kg)`/(`Height (m)`)^2)

# divide into two datasets: for all covariates and selected covariates
demo_columns_all <- c("Age","Female", "Race",
                      "Firefighter type", "BMI", 
                      "Years served as a firefighter",
                      "Non-fire-related activities (times/week)",
                      "Fire-related activities (times/month)",
                      "Current smoker","Former smoker" )
demo_columns_select <- c("Age", "BMI", 
                         "Years served as a firefighter",
                         "Non-fire-related activities (times/week)",
                         "Fire-related activities (times/month)")
demo_all <- demo %>%
  dplyr::select(1, all_of(demo_columns_all))
```

## input the missing data
```{r}
imputed_data <- mice(demo_all, m=1, method='pmm', predictorMatrix = make.predictorMatrix(demo_all))

pred <- make.predictorMatrix(demo_all)
pred[, "Non-fire-related activities (times/week)"] <- 0
pred[, "Fire-related activities (times/month)"] <- 0
pred["Non-fire-related activities (times/week)", ] <- 1
pred["Fire-related activities (times/month)", ] <- 1

imputed_data <- mice(demo_all, m=1, method='pmm', predictorMatrix = pred, seed=500)

demo_all <- complete(imputed_data, 1)

demo_select <- demo_all %>%
  dplyr::select(1, all_of(demo_columns_select))
```

# PLSDA Plot: Pre vs Post
```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#      install.packages("BiocManager")
# BiocManager::install("mixOmics")
 library(mixOmics)

# Simulate X (metabolites) and Y (group labels)

X <- metabolites_new[,5:1528]

# identify zero variance col
zero_var_cols <- which(apply(X, 2, function(col) sd(col, na.rm = TRUE) == 0))
zero_var_names <- colnames(X)[zero_var_cols]

# Remove variables with zero variance
X <- X[, apply(X, 2, function(col) sd(col, na.rm = TRUE) != 0)]

# VF vs CF
Y <- factor(metabolites_new$Group)

set.seed(123)
plsda_result <- plsda(X, Y, ncomp = 2)  # 2 components
plotIndiv(plsda_result,
          comp = c(1, 2),
          group = Y,
          legend = TRUE,
          ellipse = TRUE,
          title = "PLS-DA Score Plot")

# pre vs post
Y <- factor(metabolites_new$Time, levels = c(1, 2), labels = c("pre", "post"))

set.seed(123)
plsda_result <- plsda(X, Y, ncomp = 2)  # 2 components
plotIndiv(plsda_result,
          comp = c(1, 2),
          group = Y,
          legend = TRUE,
          ellipse = TRUE,
          title = "PLS-DA Score Plot")

```


# OPLSDA plot
```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# 
# BiocManager::install("ropls")
library(ropls)

# VF vs CF
Y <- factor(metabolites_new$Group)
set.seed(123)
opls_model <- opls(X, Y, predI = 1, orthoI = NA)  # predI = 1 component, orthoI = auto-tuned

# Extract scores
scores <- getScoreMN(opls_model)
scores_df <- as.data.frame(scores)
scores_df$Group <- Y  # Add group info

# Plot with ellipses
ggplot(scores_df, aes(x = p1, y = o1, color = Group, fill = Group)) +
  geom_point(size = 3) +
  stat_ellipse(type = "norm", alpha = 0.3) +   # << Add this line
  labs(title = "OPLS-DA Score Plot",
       x = "Predictive Component (p1)",
       y = "Orthogonal Component (o1)") +
  theme_minimal()

```



# Step 1: Paired t-test and BH correction to clean the data 
(to find metabolites that have significant difference before and after fire)

**metabolites_sig** is gained from metabolites_diff, representing the difference values of metabolites that are significant in paired t test.
**metabolites_sig_prepost** is gained from metabolites_clear, representing the pairs (post and pre values) of metabolites that are significant in paired t-test.
**metabolites_new_sig** is gained from metabolites_new that has both paired and single sample. 
*significant_CHEM* is the CHEM_ID of significant metabolites
```{r}
# Select only numeric columns for analysis
numeric_cols <- names(metabolites_clear)[sapply(metabolites_clear, is.numeric)]
numeric_cols <- setdiff(numeric_cols, "Time")  # Remove "Time" column: find all rows in numeric_cols that aren't in "Time"

# Perform paired t-test for each variable in metabolites_clear
t_test_results <- lapply(numeric_cols, function(var) {
  t_test <- t.test(pre[[var]], post[[var]], paired = TRUE)
  return(c(var, t_test$statistic, t_test$p.value))
})

# Convert list to dataframe
t_test_df <- as.data.frame(do.call(rbind, t_test_results))
colnames(t_test_df) <- c("CHEM_ID", "t_statistic", "p_value")

# Convert numeric columns to proper format
t_test_df$t_statistic <- as.numeric(t_test_df$t_statistic)
t_test_df$p_value <- as.numeric(t_test_df$p_value)

# Number of tests
num_tests <- nrow(t_test_df)

# BH-corrected p-value threshold (adjusted significance level)
alpha <- 0.05
BH_threshold <- alpha / num_tests

# Apply BH correction
t_test_df$adjusted_p_value <- p.adjust(t_test_df$p_value, method = "BH")
t_test_df$significant <- t_test_df$adjusted_p_value < alpha

# View final results with correction
print(t_test_df)

# filter the significant chemicals
t_test_clear <- t_test_df %>%
  filter(significant==TRUE)

# get the CHEM_ID of significant Chemicals
significant_CHEM <- t_test_clear$CHEM_ID
```

```{r}
# clear the metabolites
metabolites_sig <- metabolites_diff %>%
  dplyr::select(`Sample ID`, all_of(significant_CHEM))

# metabolites with pre and post
metabolites_sig_prepost <- metabolites_clear %>%
  dplyr::select(`Sample ID`, Time, Group, all_of(significant_CHEM))

# all paired and single metabolites
metabolites_new_sig <- metabolites_new %>%
  dplyr::select(PARENT_SAMPLE_NAME, `Sample ID`, Time, Group, all_of(significant_CHEM))
# write_xlsx(metabolites_sig, "metabolites_sig.xlsx")
```


# Step 2: Linear Mixed Model (find sig, same as paired t-test)
## [old code] build linear mixed models
to find if there is any difference varying by time (pre and post)
```{r}
# library(lme4)
# 
# metabolites_new$Time <- as.factor(metabolites_new$Time)
# metabolites_new$`Sample ID` <- as.factor(metabolites_new$`Sample ID`)

# Example for one metabolite (replace 'Metabolite_X' with actual column name)
# model <- lmer(`209` ~ Time + (1 | `Sample ID`), data = metabolites_new)
# summary(model)

# results <- lapply(names(metabolites_new)[5:ncol(metabolites_new)], 
#                   function(metabolite) {
#                      formula_str <- paste0("`", metabolite, "` ~ Time + (1 | `Sample ID`)")
#                       model <- lmer(as.formula(formula_str), data = metabolites_new)
#                       summary(model)
#                       })
# # example
# formula_str <- paste0("`", "212", "` ~ Time + (1 | `Sample ID`)")

```


## get the p-values etc. from the models
```{r}
# Load necessary libraries
library(lme4)
library(lmerTest)  # To obtain p-values and degrees of freedom
library(broom.mixed)  # For confidence intervals
library(openxlsx)  # To save results
library(dplyr)

metabolites_new$Time <- as.factor(metabolites_new$Time)
metabolites_new$`Sample ID` <- as.factor(metabolites_new$`Sample ID`)

# Function to fit LMM and extract statistics
results_df <- lapply(names(metabolites_new)[5:ncol(metabolites_new)], function(metabolite) {
  tryCatch({
    # Construct formula
    formula_str <- paste0("`", metabolite, "` ~ Time + (1 | `Sample ID`)")
    
    # Fit model using lmerTest::lmer() to get df
    model <- lmerTest::lmer(as.formula(formula_str), data = metabolites_new)
    
    # Print if the model is singular
    # print(isSingular(model))
    
    # Extract summary of fixed effects
    summary_model <- summary(model)
    
    # Extract coefficient matrix and convert to data frame
    fixed_effects <- as.data.frame(summary_model$coefficients)
    
    # Extract correct degrees of freedom (df) for Time effect
    df_time <- fixed_effects["Time2", "df"]  # Correct indexing
    
    # Compute confidence intervals
    ci <- confint(model, method = "Wald")
    
    # Ensure "Time2" effect exists
    time_effect_row <- which(rownames(fixed_effects) == "Time2")
    
    if (length(time_effect_row) > 0) {
      return(data.frame(
        CHEM_ID = metabolite,
        Estimate = fixed_effects["Time2", "Estimate"],
        Std_Error = fixed_effects["Time2", "Std. Error"],
        t_value = fixed_effects["Time2", "t value"],
        df = df_time,  # Correctly extracted degrees of freedom
        p_value = fixed_effects["Time2", "Pr(>|t|)"],  # Directly extract p-value
        CI_Lower = ci[time_effect_row + 1, 1],  # Lower 95% CI
        CI_Upper = ci[time_effect_row + 1, 2]   # Upper 95% CI
      ))
    } else {
      return(NULL)  # Return NULL if "Time2" effect is missing
    }
  }, error = function(e) {
    return(NULL)  # Ignore errors and continue
  })
})

# Combine results into a single data frame
results_df <- do.call(rbind, results_df)

# Mark the significance level
results_df <- results_df %>%
  mutate(
    significance_original_p = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ NA
    )
  )

# write_xlsx(results_df, "LMM_results.xlsx")
```


## do multiple comparison correction for LMM
alpha = 0.05 means that you expect 5 false positives just by chance.
Multiple testing correction controls for this inflation to ensure that your significant results are truly meaningful.
*results_sig* stores the significant metabolites with LMM.

```{r}
# Apply Bonferroni Correction
# results_df$bonferroni_p <- p.adjust(results_df$p_value, method = "bonferroni")

# Apply Benjamini-Hochberg (BH) Correction (FDR)
results_df$bh_p <- p.adjust(results_df$p_value, method = "BH")

# get the significance
results_df <- results_df %>%
  mutate(
    CHEM_ID = as.numeric(CHEM_ID),
    significance_bh_p = case_when(
      bh_p < 0.05 ~ "*",
      TRUE ~ NA
    )
  ) %>%
  left_join(chem_annotate %>% dplyr::select(CHEM_ID, PLOT_NAME), by = "CHEM_ID")


# write.xlsx(results_df, file = "LMM_results.xlsx", rowNames = FALSE)

# filter out non-significant rows
results_sig <- na.omit(results_df)

```

# Step 3: overlapped CHEM_ID with results_sig (LMM) and metabolites_sig (t-test): 
**common_CHEM_ID**

**metabo_diff_common** get the overlapped metabolites from metabolites_diff (only difference of pairs)
**metabodiff_demo** metabo_diff_common + demo: overlapped metabolites with demographic information
```{r}
CHEM_ID_t_test <- colnames(metabolites_sig)[-1]
CHEM_ID_LMM <- results_sig$CHEM_ID

# Find common elements
common_CHEM_ID <- intersect(CHEM_ID_t_test, CHEM_ID_LMM)

common_CHEM_ID

metabo_diff_common <- metabolites_diff %>%
  dplyr::select(1, all_of(common_CHEM_ID)) %>%
  mutate(`Sample ID` = gsub("FFP01", "", `Sample ID`)) %>%
  mutate(`Sample ID`= as.numeric(`Sample ID`))

metabodiff_demo <- metabo_diff_common %>%
  left_join(demo_all, metabo_diff_common, by = "Sample ID")

```

## Chemical Annotation sheet
```{r}
chem_annotate <- read_excel("for_code_UNTX-07-23PHML+ DATA TABLES.xlsx", sheet = "Chemical Annotation")

filtered_data <- chem_annotate %>% filter(CHEM_ID %in% common_CHEM_ID)
t_test_metabo <- chem_annotate %>% filter(CHEM_ID %in% CHEM_ID_t_test)
LMM_metabo <- chem_annotate %>% filter(CHEM_ID %in% CHEM_ID_LMM)

# write.xlsx(filtered_data, "22_sig_metabo_annotation.xlsx")
# write.xlsx(t_test_metabo, "t_test_metabo_annotation.xlsx")
# write.xlsx(LMM_metabo, "LMM_metabo_annotation.xlsx")


```

# Volcano plots
```{r}
# 1. Prepare your data
fc_pos  <- log2(1.5)    # ≈ 0.585
fc_neg  <- -log2(1.5)   # ≈ −0.585

volcano_df <- results_df %>%
  mutate(
    logFC     = Estimate,
    negLog10P = -log10(bh_p),
    status = case_when(
      bh_p < 0.05 & logFC >= fc_pos  ~ "Up (>|1.5|)",
      bh_p < 0.05 & logFC <= fc_neg  ~ "Down (>|1.5|)",
      TRUE                           ~ "NS"
    )
  ) 


# 2. Plot
ggplot(volcano_df, aes(x = logFC, y = negLog10P, color = status)) +
  geom_point(alpha = 0.7, size = 2) +
  # horizontal line at adj-p = 0.05
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey50") +
  # vertical lines at ±log2(1.5)
  geom_vline(xintercept = c(fc_neg, fc_pos), linetype = "dashed", color = "grey50") +
  scale_color_manual(
    values = c(
      "Up (>|1.5|)"   = "red",
      "Down (>|1.5|)" = "blue",
      "NS"            = "grey30"
    )
  ) +
  labs(
    x     = expression(log[2]~Fold~Change~"(Estimate)"),
    y     = expression(-log[10]~"(BH‐adjusted p)"),
    color = NULL,
    title = "Volcano Plot of LMM Results\n(FC > 1.5 & BH p < 0.05)"
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "right") +
  # 3. label only the hit points
  geom_text_repel(
    data = subset(volcano_df, status != "NS"),
    aes(label = PLOT_NAME),
    size        = 3,
    max.overlaps = 20
  )
```



# Spagetti Plot

### all 98 significant metabolites
```{r}

# Step 1: Reshape to long format
met_long <- metabolites_sig_prepost %>%
  pivot_longer(cols = 4:ncol(.), names_to = "ChemID", values_to = "Value")

# Step 2: Prepare plot data with original Time and Group info
plot_data_long <- met_long %>%
  rename(SampleID = `Sample ID`) %>%
  mutate(Time = ifelse(Time == 1, "Pre", "Post"),
         Time = factor(Time, levels = c("Pre", "Post")))

# Step 3: Compute group-level and overall means (still using raw values)
mean_lines <- plot_data_long %>%
  group_by(Time, Group) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE), .groups = "drop")

overall_mean <- plot_data_long %>%
  group_by(Time) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE))

# Step 4: Plot
ggplot(plot_data_long, aes(x = Time, y = Value, group = interaction(SampleID, ChemID))) +
  geom_line(color = "gray", alpha = 0.4) +
  geom_line(data = overall_mean, aes(y = mean_value, group = 1),
            color = "black", size = 1.2) +
  geom_line(data = mean_lines %>% filter(Group == "CF"),
            aes(y = mean_value, group = Group),
            color = "red", linetype = "dashed", size = 1) +
  geom_line(data = mean_lines %>% filter(Group == "VF"),
            aes(y = mean_value, group = Group),
            color = "blue", linetype = "dashed", size = 1) +
  labs(x = "Time", y = "Metabolite Value", title = "Metabolite Levels: Pre vs. Post by Group") +
  theme_minimal()

```

### VF and CF separately
```{r}

library(tidyverse)

# Step 1: Reshape to long format
met_long <- metabolites_sig_prepost %>%
  pivot_longer(cols = 4:ncol(.), names_to = "ChemID", values_to = "Value")

# Step 2: Prepare full plot data
plot_data_long <- met_long %>%
  rename(SampleID = `Sample ID`) %>%
  mutate(Time = ifelse(Time == 1, "Pre", "Post"),
         Time = factor(Time, levels = c("Pre", "Post")))

# Now split into two groups
cf_data <- plot_data_long %>% filter(Group == "CF")
vf_data <- plot_data_long %>% filter(Group == "VF")

# Step 3a: CF Plot
cf_mean <- cf_data %>%
  group_by(Time) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE))

ggplot(cf_data, aes(x = Time, y = Value, group = interaction(SampleID, ChemID))) +
  geom_line(color = "gray", alpha = 0.4) +
  geom_line(data = cf_mean, aes(y = mean_value, group = 1),
            color = "red", linewidth = 1.2) +
  labs(x = "Time", y = "Metabolite Value", title = "CF Group: Pre vs. Post Metabolite Levels") +
  theme_minimal()

# Step 3b: VF Plot
vf_mean <- vf_data %>%
  group_by(Time) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE))

ggplot(vf_data, aes(x = Time, y = Value, group = interaction(SampleID, ChemID))) +
  geom_line(color = "gray", alpha = 0.4) +
  geom_line(data = vf_mean, aes(y = mean_value, group = 1),
            color = "blue", linewidth = 1.2) +
  labs(x = "Time", y = "Metabolite Value", title = "VF Group: Pre vs. Post Metabolite Levels") +
  theme_minimal()

```

### mean values for 98 significant metabolites
```{r}
library(tidyverse)

# Step 1: reshape to long
met_long <- metabolites_sig_prepost %>%
  pivot_longer(cols = 4:ncol(.),
               names_to  = "ChemID",
               values_to = "Value")

# Step 2: add nice Time labels
plot_data_long <- met_long %>%
  rename(SampleID = `Sample ID`) %>%
  mutate(Time = factor(ifelse(Time == 1, "Pre", "Post"),
                       levels = c("Pre","Post")))

# Step 3: compute per‐metabolite means at Pre and Post
metabolite_means <- plot_data_long %>%
  group_by(ChemID, Time) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE),
            .groups = "drop")

# Step 4: compute group‐level and overall means
group_means <- plot_data_long %>%
  group_by(Time, Group) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE),
            .groups = "drop")

overall_means <- plot_data_long %>%
  group_by(Time) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE),
            .groups = "drop")

# Step 5: draw the spaghetti plot
ggplot() +
  # 98 grey lines (one per metabolite)
  geom_line(data = metabolite_means,
            aes(x = Time, y = mean_value, group = ChemID),
            color = "gray", alpha = 0.6) +
  # overall mean
  geom_line(data = overall_means,
            aes(x = Time, y = mean_value, group = 1),
            color = "black", size = 1.2) +
  # CF mean
  geom_line(data = filter(group_means, Group == "CF"),
            aes(x = Time, y = mean_value, group = Group),
            color = "red", linetype = "dashed", size = 1) +
  # VF mean
  geom_line(data = filter(group_means, Group == "VF"),
            aes(x = Time, y = mean_value, group = Group),
            color = "blue", linetype = "dashed", size = 1) +
  labs(
    x     = "Time",
    y     = "Mean metabolite value",
    title = "Each metabolite’s Pre→Post change, with group & overall means"
  ) +
  theme_minimal()

```



## Venn Diagram
```{r}
library(VennDiagram)
# Example data
set_t_test <- CHEM_ID_t_test
set_LMM <- CHEM_ID_LMM

# Generate Venn diagram
venn.plot <- venn.diagram(
  x = list(Set1 = set_t_test, Set2 = set_LMM),
  category.names = c("set_t_test", "set_LMM"),
  filename = NULL,
  fill = c("red", "blue") # Colors for each set
)

# Display the Venn diagram
grid.draw(venn.plot)
```


# Step 4: correlation coefficient table
in metabodiff_demo, all firefighters are white, only 2 of them are volunteers, none of them smoking, only one female.
Figure out which covariates are significantly correlated to the metabolites.
**metaodiff_demo_clear** is cleared metabodiff_demo w/o race, CF/VF, smoke, female; and have one data imputed in "Years served as a firefighter"
only "Fire-related activities (times/month)" has two non-significant p-values
```{r}
metabodiff_demo_select <- metabodiff_demo %>% 
  dplyr::select(all_of(c("Sample ID",common_CHEM_ID, demo_columns_select)))

# Generate the correlation coefficient table
correlation_table <- metabodiff_demo_select %>%
  cor(method = "spearman") %>%
  .[common_CHEM_ID, demo_columns_select]

# t-test and p-values
get_p_values <- function(df, common_CHEM_ID, demo_columns_select) {
  p_values <- matrix(NA, nrow = length(common_CHEM_ID), ncol = length(demo_columns_select))
  rownames(p_values) <- common_CHEM_ID
  colnames(p_values) <- demo_columns_select
  
  for (chem_id in common_CHEM_ID) {
    for (demo_col in demo_columns_select) {
      group1 <- metabodiff_demo_select[[chem_id]]
      group2 <- metabodiff_demo_select[[demo_col]]
      t_test_result <- t.test(group1, group2)
      p_values[chem_id, demo_col] <- t_test_result$p.value
    }
  }
  return(as.data.frame(p_values))
}

# Get p-values for the correlation table
p_values_table <- get_p_values(metabodiff_demo_select, common_CHEM_ID, demo_columns_select)

# write.xlsx(p_values_table, file = "corr_p_values_table.xlsx", rowNames = FALSE)
```


# Step 5: 2nd model: LMMs with all covariates: 
Based on the result we got from the above, we have metabolites that have significant difference, common_CHEM_ID, then extract the common_CHEM_ID from the metabolites_new.
```{r}

# combine the demo to metabolites_new (pairs and single)
metabolites_new_common <- metabolites_new %>%
  dplyr::select(1:4, all_of(common_CHEM_ID)) %>%
  mutate(`Sample ID` = gsub("^([0-9]{1,2}).*", "\\1", `Sample ID`)) %>%
  mutate(`Sample ID` = as.numeric(`Sample ID`))

metabonew_demo <- metabolites_new_common %>%
  left_join(demo_all, metabolites_new_common, by = "Sample ID")

# just remove the first PARENT_SAMPLE_NAME column
metabonew_demo_clear <- metabonew_demo[,-1]

# Function to fit LMM and extract statistics
results_df_2 <- lapply(common_CHEM_ID, function(metabolite) {
  tryCatch({
    # Construct formula with additional covariates
    formula_str <- paste0("`", metabolite, "` ~ Time + Age + Female + Race + BMI + `Firefighter type` + `Years served as a firefighter` + `Non-fire-related activities (times/week)` + `Fire-related activities (times/month)` + `Current smoker` + `Former smoker` + (1 | `Sample ID`)")
    
    # Fit model using lmerTest::lmer() to get df
    model <- lmerTest::lmer(as.formula(formula_str), data = metabonew_demo_clear)
    
    # Extract summary of fixed effects
    summary_model <- summary(model)
    
    # Extract coefficient matrix and convert to data frame
    fixed_effects <- as.data.frame(summary_model$coefficients)
    
    # Compute confidence intervals
    ci <- confint(model, method = "Wald")
    
    # Extract statistics for each covariate
    covariates <- rownames(fixed_effects)[-1]
    results_list <- lapply(covariates, function(covariate) {
      if (covariate %in% rownames(fixed_effects)) {
        return(data.frame(
          CHEM_ID = metabolite,
          Covariate = covariate,
          Estimate = fixed_effects[covariate, "Estimate"],
          Std_Error = fixed_effects[covariate, "Std. Error"],
          t_value = fixed_effects[covariate, "t value"],
          df = fixed_effects[covariate, "df"],
          p_value = fixed_effects[covariate, "Pr(>|t|)"],
          CI_Lower = ci[covariate, 1],
          CI_Upper = ci[covariate, 2]
        ))
      } else {
        return(NULL)
      }
    })
    
    # Combine results for all covariates
    do.call(rbind, results_list)
  }, error = function(e) {
    return(NULL)  # Ignore errors and continue
  })
})

# Combine results into a single data frame
results_df_2 <- do.call(rbind, results_df_2)

# Mark the significance level
results_df_2 <- results_df_2 %>%
  mutate(
    significance_original_p = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ NA
    )
  )

# filter out non-significant rows
results_sig_2 <- na.omit(results_df_2)
```
## 2nd model: LMMs with selected covariates: 

```{r}
# Function to fit LMM and extract statistics
results_df_3 <- lapply(common_CHEM_ID, function(metabolite) {
  tryCatch({
    # Construct formula with additional covariates
    formula_str <- paste0("`", metabolite, "` ~ Time + Age + BMI + `Years served as a firefighter` + `Non-fire-related activities (times/week)` + `Fire-related activities (times/month)` + (1 | `Sample ID`)")
    
    # Fit model using lmerTest::lmer() to get df
    model <- lmerTest::lmer(as.formula(formula_str), data = metabonew_demo_clear)
    
    # Extract summary of fixed effects
    summary_model <- summary(model)
    
    # Extract coefficient matrix and convert to data frame
    fixed_effects <- as.data.frame(summary_model$coefficients)
    
    # Compute confidence intervals
    ci <- confint(model, method = "Wald")
    
    # Extract statistics for each covariate
    covariates <- rownames(fixed_effects)[-1]
    results_list <- lapply(covariates, function(covariate) {
      if (covariate %in% rownames(fixed_effects)) {
        return(data.frame(
          CHEM_ID = metabolite,
          Covariate = covariate,
          Estimate = fixed_effects[covariate, "Estimate"],
          Std_Error = fixed_effects[covariate, "Std. Error"],
          t_value = fixed_effects[covariate, "t value"],
          df = fixed_effects[covariate, "df"],
          p_value = fixed_effects[covariate, "Pr(>|t|)"],
          CI_Lower = ci[covariate, 1],
          CI_Upper = ci[covariate, 2]
        ))
      } else {
        return(NULL)
      }
    })
    
    # Combine results for all covariates
    do.call(rbind, results_list)
  }, error = function(e) {
    return(NULL)  # Ignore errors and continue
  })
})

# Combine results into a single data frame
results_df_3 <- do.call(rbind, results_df_3)

# Mark the significance level
results_df_3 <- results_df_3 %>%
  mutate(
    significance_original_p = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ NA
    )
  )

# filter out non-significant rows
results_sig_3 <- na.omit(results_df_3)

```

# Pathway Analysis: extract the overlapped CHEM_IDs from Chemical Annotation sheet

```{r}
pa_results <- read_excel("PA_Results_0507.xlsx")
PA_metabo_names <- c("L-Glutamate", "L-Arginine", "L-Aspartate", "2-Oxoglutarate", "Fumarate", "L-Aspartic acid", "Oxoglutaric acid", "D-Glutamic acid", "L-Malic acid", "Pyruvic acid", "Carnosine", "Uracil", "Pyruvate", "Cys-Gly", "5-Oxoproline", "Spermidine", "Spermine", "(S)-Malate", "(S)-Lactate", "Sphingosine 1-phosphate", "Sphinganine 1-phosphate", "Sphinganine", "Sphingosine")
LMM_metabo_names <- LMM_metabo$CHEMICAL_NAME
common_PA_LMM <- intersect(PA_metabo_names, LMM_metabo_names)
```



